On the Bit Complexity of Finding Points in Connected Components of a Smooth Real Hypersurface

### 2020-02-07 main.tex
Degree bounds

This being said, our two key results are the following.
\begin{theorem}\label{theo:gen}
  For $i=1,\dots,n$, there exists a non-zero polynomial $\D_i$ in
  $\C[\A]$ of degree at most $\td$ such that if $\mA \in \C^{n\times
    n}$ does not cancel $\D_i$, then $\mA$ is invertible and $f^\mA$
  satisfies $\textbf{H}_i$.
\end{theorem}

\begin{theorem}
  Suppose that $f$ satisfies $\textbf{H}_i$. There exists a non-zero
  polynomial $\Xi_{i} \in \C[S_1,\dots,S_{i-1}]$ of degree at most
  $d^{2n}$ such that if $\bm \sigma \in \C^{i-1}$ does not
  cancel $\Xi_{i}$, then $f$ and $\bm \sigma$ satisfy $\textbf{H}_i'$.
\end{theorem}
\noindent 
The proof of these theorems occupies the next two sections.

@@

Let $\bm\Psi: \C^n \times \C^{i-1} \rightarrow \C^{n}$ be the mapping defined by the polynomials
\[
  (X_1-S_1,\hdots,X_{i-1}-S_{i-1}, f,\frac{\pa f}{\pa X_{i+1}},\hdots,\frac{\pa f}{\pa X_n} ).
\]
%
\begin{lemma}
  $\bz$ is a regular value of $\bm\Psi.$
\end{lemma}
\begin{proof}
At all $(\xb,\bm\sigma)$ zeros of $\bm\Psi,$ the Jacobian matrix of
$\bm\Psi$ has full rank $n$. Indeed, indexing columns 
by $X_1,\dots,X_n,S_1,\dots,S_{i-1}$, this matrix is equal to
\[
\left[ 
\begin{array}{cc}
\bI_{i-1}\hspace{7mm} \bz_{(i-1)\times (n-i+1)}&\hspace{-10mm}-\bI_{i-1}\\
\jac_{\xb}\left(f,\frac{\pa
  f}{\pa X_{i+1}},\hdots,\frac{\pa f}{\pa X_n}\right)&\bz_{(n-i+1)\times (i-1)}
\end{array}
\right].
\] 
Since the Jacobian of $f,\frac{\pa f}{\pa X_{i+1}},\hdots,\frac{\pa
  f}{\pa X_n}$ at $\xb$ is non-zero (by $\textbf{H}_i$), the entire
matrix must have full rank $n$. Thus,  $\bz$ is a regular value of $\bm\Phi.$
\end{proof}
%
Since all polynomials defining $\bm\Psi$ have degree at most $d$, it
follows by Proposition~\ref{prop:weak_t} that there exists a non-zero
polynomial $\Xi_i$ in $\C[S_1,\dots,S_{i-1}]$ of degree at most
$d^{2n},$ with the following property: at any root $\xb$ of
$(X_1-\sigma_1,\dots,X_{i-1}-\sigma_{i-1},f, {\pa f}/{\pa
  X_{i+1}},\dots,{\pa f}/{\pa X_n})$, the Jacobian matrix of these
equations has rank $n$. The conclusion follows.

@@

Let us then define 
$$\D_{i}:=\D_{1,i} \alpha_i \cdots \alpha_n D_i \cdots D_n,$$
where, for all $j$, $\alpha_j$ is as above and $D_j$ is the leading
coefficient of $\pjb$ with respect to $X_j$. Thus, $\D_{i}$ is a
non-zero polynomial in $\C[\A]$; we will estimate its degree below.

\begin{lemma}
  Suppose that $\mA \in \C^{n\times n}$ does not cancel
  $\D_{i}$. Then $f^\mA$ satisfies ${\bf H}_i$.
\end{lemma}
\begin{proof}
  By assumption, $\D_{2,i}(\mA)$ is non-zero, so that $\mA$ is
  invertible and $f^\mA$ satisfies ${\bf H}_i(1)$. In particular, the
  ideal $\I(i, f^\mA)$ is radical, and its zero-set
  $W(\pi_i,V(f^\mA))$ is either empty or $(i-1)$-equidimensional. If it is 
  empty, we are done.

@@

\td finish the degree bound.

% \section{Proof of Theorem 2.1}
% %
% Recall Theorem 3.1: \textit{Suppose that $f \in \Q[X_1,\hdots,X_n]$ satisfies \textbf{H}. Then there exists a Zariski closed set $\D \subset \mathbb{C}^{n^2}$ of degree at most $D^{\textrm{\td}}$ such that, if $\textbf{A}$ is chosen from  $\C^{n^2}-\Delta$ ( $\Q^{n^2}-\Delta$?) then $f^A$ satisfies $\textbf{H}_i^{'}$ for all $i \in \{1,\hdots,n\}.$} 
% %
% \begin{proof}
% Assume that $f\in \Q[X_1,\hdots,X_n]$ has degree at most $D$ and satisfies $\textbf{H}.$ By Theorem 5.1, there exists a Zariski closed set $\D_1\subset \C^{n^2}$ with the property that, if $A \in \C^{n^2}-\D_1$ then, for $i \in \{1,\hdots,n\},$ $f^A$ satisfies $\textbf{H}_i^{'}(1), \textbf{H}_i^{'}(2),$ and $\textbf{H}_i^{'}(3).$ Likewise, by Theorem 6.1, we obtain $\D_2 \subset \C^{n^2}$ with the property that if $A \in \C^{n^2}-\D_2$ then $f^A$ satisfies $\textbf{H}_i^{'}(4), \textbf{H}_i^{'}(5),$ and $\textbf{H}_i^{'}(6).$ By Theorem 7.1, $\D_3 \subset \C^{n^2}$ exists with $f^A$ satisfying $\textbf{H}_i^{'}(7)$ when $A \in \C^{n^2}-\D_3$. Clearly then, if 
% \[
% A \in \C^{n^2} - \bigcup_{i=1}^3 \D_i
% \]
% then $f^A$ satisfies  $\textbf{H}_i^{'}$ for all $i \in \{1,\hdots,n\}.$ Take 
% \[
% \Delta = \bigcup_{i=1}^3 \D_i,
% \]
% and note that 
% \[
% \deg \D \leq  \sum_{i=1}^3 \deg \D_i.
% \]
% \end{proof}
% %
% Recall Corollary 3.2: \textit{Fix $S \subset \mathbb{Q}$ with $|S| \geq \epsilon^{-1} D^{\textrm{\td}}$ and $\epsilon > 0$. Then for $A\in S^{n^2}$ chosen randomly, the probability that $f^A$ satisfies $H_i^{'}$ for all $i \in \{1,\hdots,n\}$ is at least $1-\epsilon.$}
% %
% \begin{proof}
% \[
% Pr[A \in \Delta] \leq \frac{\deg \Delta}{|S|} \leq \frac{D^{\td}}{\epsilon^{-1}D^{\td}} = \epsilon.
% \]
% \end{proof}
% %
% %
% %
% %
\section{Bit complexity}
Recall Theorem 1.1:   Suppose that $f\in\ZZ[X_1\hdots,X_n]$ is squarefree, satisfies
  $\deg(f) \leq d, \htt(f) \leq s$, and that $V(f) \subset \C^n$ is
  smooth.  Assume also that $f$ is given by a straight-line program
  $\Gamma$ of size $L$ with integer constants of height at most $b.$

@@

  With probability at least $1-\epsilon$, the output is correct. Otherwise, the algorithm either returns a proper subset of the points or FAIL. In any case, the algorithm uses
  \[
  O^{\sim}(nk(Lb+d^{2n}(s + \log 1/\epsilon+d)(L+d)\log(s + \log 1/\epsilon)))
  \]
  bit operations, with $k \geq 1+\lg n + \lg(1/\epsilon)$. The polynomials in
  the output have degree at most $d^n,$ and height 
  \[
  O^{\sim}((s+\lg 1/\epsilon + d)d^n)
  =O^{\sim}((s+\lg 1/\epsilon)d^{n+1}).
  \]
%
\subsection{The Algorithm}
%
For a matrix $A \in \GL(n,\Q),$ let 
\[
Y_i = X_i^{A} = [b_i] \cdot \bbm X_1\\ \vdots \\ X_n \ebm, 
\]
where $b_i$ is the i-th row of $A^{-1}$.
%
\begin{algorithm}[hbt!]
\KwIn{a strait line program $\Gamma$ that computes $f \in \Q[X_1,\hdots,X_n],$ and $d=\deg(f)$}
\KwOut{$n$ zero-dimensional parameterizations, the union of which includes at least one point in each connected component of $V(f) \cap \R^n$}
\nl Take  $S := \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{3n}\}$
 and randomly choose $A \in S^{n^2}$, and  $\sigma_{n-1}=(\sigma_1,\hdots,\sigma_{n-1}) \in S^{n-1}$\;
\nl Compute $f$ and apply the change of variables to obtain $f(A x)=f^A(x)$\;
\nl Use \cite{partialDerComputation} to compute the partial derivatives $\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n}$\;
    \caption{{Main Algorithm} \label{}}
\nl \For{$i\gets1$ \KwTo $n$}{
\nl Run \cite[Algorithm 2]{SH} $k \geq 1+\lg n + \lg(1/\epsilon)$ times with input (state this in terms of straight line programs?)
$\{Y_1-\sigma_1,\hdots,Y_{i-1}-\sigma_{i-1},f^A,\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n} \} \textrm{ and } d$\;
\nl Let $\mathscr{Q}_i$ be the highest cardinality zero-dimensional parameterization returned in step 5 \;
} 
\nl \Return $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$.
\end{algorithm}
\clearpage
%
\subsection{Proof of Theorem 1.1}
%
\textbf{Bit operation cost:} The following lists the costs for each step of Algorithm 1:
%
\begin{enumerate}
    \item We defined $S = \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{3n}\}$ and therefore the height of $a_{i,j} \in S$ is at most 
    \[
    \log 1/{\epsilon} + \log(3n^2(2d)^{2n}) = O^{\sim}(\log 1/{\epsilon}).
    \]
    \item 
    \begin{enumerate}[label=\roman*.]
        \item $O(L)$ operations are required to compute $f$ from its straight line program $\Gamma.$
        \item After applying the change of variables obtaining $f(A x)=f^A(x)$ the height grows at most by a factor of
        \begin{align*}
        \log(n!\cdot \max_{\frac{1\leq i\leq n}{1\leq j\leq n}}\{a_{ij}\}) &= 
        O^{\sim}(n\log n + \log 1/{\epsilon})\\
        &= O^{\sim}(\log 1/{\epsilon}).
        \end{align*}
    \end{enumerate}
    \item After computing the partial derivatives the height grows by at most another factor of $\log d$. Furthermore, preparing the straight line program for the gradient of $f$ does not introduce large integers. The integers in $\grad (f)$ are of the same magnitude as in $\Gamma$.
    \item 
    After applying \cite[Algorithm 2]{SH} $kn$ times, we obtain the
    following upper bound on the total number of bit operations required by the algorithm:
  \[
  O^{\sim}(nk(Lb+d^{2n}(s + \log 1/\epsilon+d)(L+d)\log(s + \log 1/\epsilon)))
  \]
    where it also follows from \cite[Algorithm 2]{SH} that the polynomials in the output have degree at most $d^n,$ and height at most
  \[
  O^{\sim}((s+\lg 1/\epsilon + d)d^n)
  =O^{\sim}((s+\lg 1/\epsilon)d^{n+1}).
  \]
\end{enumerate}
%
\par \noindent 
\textbf{Probability of success:} By Theorem 2.1, there exists $\Delta_i \in \C[\A]$ with $\deg \Delta_i \leq 2n^2(2d)^{3n}.$ By Theorem 2.2, there exists $\Xi_i \subset \C[\A]$ with $\deg \Xi_i \leq d^{2n}.$ Take $\Delta =  \prod_{i=1}^n\Delta_i  \Xi_i $ so that 
%
\begin{align*}
    \deg \Delta &\leq \sum_{i=1}^n (\deg \Delta_i + \deg \Xi_i)\\
    &\leq 2n^2(2d)^{3n} + nd^{2n}\\
    &\leq 3n^2(2d)^{3n}, 
\end{align*}
%
and note that, for $A \in \C^{n \times n}$ that does not cancel $\Delta,$  $\mA$ is then invertible and $f^\mA$ satisfies $\textbf{H}_i$ and $\textbf{H}_i^{'}$ for all $i \in \{1,\hdots,n\}.$

>>>>

This being said, our two key results are the following.
\begin{theorem}\label{theo:gen}
  For $i=1,\dots,n$, there exists a non-zero polynomial $\D_i$ in
  $\C[\A]$ of degree at most $6n^2(2d)^{2n}$ such that if $\mA \in
  \C^{n\times n}$ does not cancel $\D_i$, then $\mA$ is invertible and
  $f^\mA$ satisfies $\textbf{H}_i$.
\end{theorem}

\begin{theorem}\label{theo:Hp}
  Suppose that $f$ satisfies $\textbf{H}_i$. There exists a non-zero
  polynomial $\Xi_{i} \in \C[S_1,\dots,S_{i-1}]$ of degree at most
  $d^{2n}$ such that if $\bm \sigma \in \C^{i-1}$ does not
  cancel $\Xi_{i}$, then $f$ and $\bm \sigma$ satisfy $\textbf{H}_i'$.
\end{theorem}
\noindent 
The proof of these theorems occupies the next two sections.

@@

Let $\bm\Psi: \C^n \times \C^{i-1} \rightarrow \C^{n}$ be the mapping defined by the polynomials
\[
  (X_1-S_1,\hdots,X_{i-1}-S_{i-1}, f,\frac{\pa f}{\pa X_{i+1}},\hdots,\frac{\pa f}{\pa X_n} ).
\]
%
\begin{lemma}
  $\bz$ is a regular value of $\bm\Psi.$
\end{lemma}
\begin{proof}
At all $(\xb,\bm\sigma)$ zeros of $\bm\Psi,$ the Jacobian matrix of
$\bm\Psi$ has full rank $n$. Indeed, indexing columns 
by $X_1,\dots,X_n,S_1,\dots,S_{i-1}$, this matrix is equal to
\[
\left[ 
\begin{array}{cc}
\bI_{i-1}\hspace{7mm} \bz_{(i-1)\times (n-i+1)}&\hspace{-10mm}-\bI_{i-1}\\
\jac_{\xb}\left(f,\frac{\pa
  f}{\pa X_{i+1}},\hdots,\frac{\pa f}{\pa X_n}\right)&\bz_{(n-i+1)\times (i-1)}
\end{array}
\right].
\] 
Since the Jacobian of $f,\frac{\pa f}{\pa X_{i+1}},\hdots,\frac{\pa
  f}{\pa X_n}$ at $\xb$ is non-zero (by $\textbf{H}_i$), the entire
matrix must have full rank $n$. Thus,  $\bz$ is a regular value of $\bm\Phi.$
\end{proof}

Since all polynomials defining $\bm\Psi$ have degree at most $d$, it
follows by Proposition~\ref{prop:weak_t} that there exists a non-zero
polynomial $\Xi_i$ in $\C[S_1,\dots,S_{i-1}]$ of degree at most
$d^{2n},$ with the following property: at any root $\xb$ of
$(X_1-\sigma_1,\dots,X_{i-1}-\sigma_{i-1},f, {\pa f}/{\pa
  X_{i+1}},\dots,{\pa f}/{\pa X_n})$, the Jacobian matrix of these
equations has rank $n$. Theorem~\ref{theo:Hp} is proved.

@@

Let us then define 
$$\D_{i}:=\D_{i,1} \alpha_i \cdots \alpha_n D_i \cdots D_n,$$ where
$\D_{i,1}$ was defined in Section~\ref{ssec:Hi1} and for all $j$,
$\alpha_j$ is as above and $D_j$ is the leading coefficient of $\pjb$
with respect to $X_j$. Thus, $\D_{i}$ is a non-zero polynomial in
$\C[\A]$; we will estimate its degree below.

\begin{lemma}
  Suppose that $\mA \in \C^{n\times n}$ does not cancel $\D_{i}$. Then
  $f^\mA$ satisfies ${\bf H}_i$.
\end{lemma}
\begin{proof}
  By assumption, $\D_{i,1}(\mA)$ is non-zero, so that $\mA$ is
  invertible and $f^\mA$ satisfies ${\bf H}_i(1)$. In particular, the
  ideal $\I(i, f^\mA)$ is radical, and its zero-set
  $W(\pi_i,V(f^\mA))$ is either empty or $(i-1)$-equidimensional. If
  it is empty, we are done.

@@

We saw in Section~\ref{ssec:Hi1} that $\D_{i,1}$ has degree at most
$2nd^{2n}$, and all $D_j$'s have degree at most $(2d)^n$. This gives
the upper bound
$$\deg(\D_i) \le 2nd^{2n} + 2n^2(2d)^{2n} + n(2d)^n \le 6n^2 (2d)^{2n}.$$
This finishes the proof of Theorem~\ref{theo:gen}.

\section{Bit complexity}
Recall Theorem 1.1:   Suppose that $f\in\ZZ[X_1\hdots,X_n]$ is squarefree, satisfies
  $\deg(f) \leq d, \htt(f) \leq s$, and that $V(f) \subset \C^n$ is
  smooth.  Assume also that $f$ is given by a straight-line program
  $\Gamma$ of size $L$ with integer constants of height at most $b.$

@@

  With probability at least $1-\epsilon$, the output is correct. Otherwise, the algorithm either returns a proper subset of the points or FAIL. In any case, the algorithm uses
  \[
  O^{\sim}(nk(Lb+d^{2n}(s + \log 1/\epsilon+d)(L+d)\log(s + \log 1/\epsilon)))
  \]
  bit operations, with $k \geq 1+\lg n + \lg(1/\epsilon)$. The polynomials in
  the output have degree at most $d^n,$ and height 
  \[
  O^{\sim}((s+\lg 1/\epsilon + d)d^n)
  =O^{\sim}((s+\lg 1/\epsilon)d^{n+1}).
  \]
%
\subsection{The Algorithm}
%
For a matrix $A \in \GL(n,\Q),$ let 
\[
Y_i = X_i^{A} = [b_i] \cdot \bbm X_1\\ \vdots \\ X_n \ebm, 
\]
where $b_i$ is the i-th row of $A^{-1}$.
%
\begin{algorithm}[hbt!]
\KwIn{a strait line program $\Gamma$ that computes $f \in \Q[X_1,\hdots,X_n],$ and $d=\deg(f)$}
\KwOut{$n$ zero-dimensional parameterizations, the union of which includes at least one point in each connected component of $V(f) \cap \R^n$}
\nl Take  $S := \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{3n}\}$
 and randomly choose $A \in S^{n^2}$, and  $\sigma_{n-1}=(\sigma_1,\hdots,\sigma_{n-1}) \in S^{n-1}$\;
\nl Compute $f$ and apply the change of variables to obtain $f(A x)=f^A(x)$\;
\nl Use \cite{partialDerComputation} to compute the partial derivatives $\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n}$\;
    \caption{{Main Algorithm} \label{}}
\nl \For{$i\gets1$ \KwTo $n$}{
\nl Run \cite[Algorithm 2]{SH} $k \geq 1+\lg n + \lg(1/\epsilon)$ times with input (state this in terms of straight line programs?)
$\{Y_1-\sigma_1,\hdots,Y_{i-1}-\sigma_{i-1},f^A,\frac{\pa f^A}{\pa X_{i+1}},\hdots,\frac{\pa f^A}{\pa X_n} \} \textrm{ and } d$\;
\nl Let $\mathscr{Q}_i$ be the highest cardinality zero-dimensional parameterization returned in step 5 \;
} 
\nl \Return $[\mathscr{Q}_1,\hdots,\mathscr{Q}_n]$.
\end{algorithm}

%
\subsection{Proof of Theorem 1.1}
%
\textbf{Bit operation cost:} The following lists the costs for each step of Algorithm 1:
%
\begin{enumerate}
    \item We defined $S = \{1,2,\hdots,2\epsilon^{-1}3n^2(2d)^{3n}\}$ and therefore the height of $a_{i,j} \in S$ is at most 
    \[
    \log 1/{\epsilon} + \log(3n^2(2d)^{2n}) = O^{\sim}(\log 1/{\epsilon}).
    \]
    \item 
    \begin{enumerate}[label=\roman*.]
        \item $O(L)$ operations are required to compute $f$ from its straight line program $\Gamma.$
        \item After applying the change of variables obtaining $f(A x)=f^A(x)$ the height grows at most by a factor of
        \begin{align*}
        \log(n!\cdot \max_{\frac{1\leq i\leq n}{1\leq j\leq n}}\{a_{ij}\}) &= 
        O^{\sim}(n\log n + \log 1/{\epsilon})\\
        &= O^{\sim}(\log 1/{\epsilon}).
        \end{align*}
    \end{enumerate}
    \item After computing the partial derivatives the height grows by at most another factor of $\log d$. Furthermore, preparing the straight line program for the gradient of $f$ does not introduce large integers. The integers in $\grad (f)$ are of the same magnitude as in $\Gamma$.
    \item 
    After applying \cite[Algorithm 2]{SH} $kn$ times, we obtain the
    following upper bound on the total number of bit operations required by the algorithm:
  \[
  O^{\sim}(nk(Lb+d^{2n}(s + \log 1/\epsilon+d)(L+d)\log(s + \log 1/\epsilon)))
  \]
    where it also follows from \cite[Algorithm 2]{SH} that the polynomials in the output have degree at most $d^n,$ and height at most
  \[
  O^{\sim}((s+\lg 1/\epsilon + d)d^n)
  =O^{\sim}((s+\lg 1/\epsilon)d^{n+1}).
  \]
\end{enumerate}
%
\par \noindent 
\textbf{Probability of success:} By Theorem 2.1, there exists $\Delta_i \in \C[\A]$ with $\deg \Delta_i \leq 2n^2(2d)^{3n}.$ By Theorem 2.2, there exists $\Xi_i \subset \C[\A]$ with $\deg \Xi_i \leq d^{2n}.$ Take $\Delta =  \prod_{i=1}^n\Delta_i  \Xi_i $ so that 
%
\begin{align*}
    \deg \Delta &\leq \sum_{i=1}^n (\deg \Delta_i + \deg \Xi_i)\\
    &\leq 2n^2(2d)^{3n} + nd^{2n}\\
    &\leq 3n^2(2d)^{3n}, 
\end{align*}
%
and note that, for $A \in \C^{n \times n}$ that does not cancel $\Delta,$  $\mA$ is then invertible and $f^\mA$ satisfies $\textbf{H}_i$ and $\textbf{H}_i^{'}$ for all $i \in \{1,\hdots,n\}.$
